---
layout: post
title: "Calculating  Risk"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

I was playing a game of Risk a few days ago with my family. As any Risk player knows, the game is almost entirely luck-based. The dice you roll and the cards you draw matter far more than any strategic decision you make. This reality, coupled with the fact that most people get far more emotionally involved than would seem warranted, means that many people develop their own methods of beating the odds. These methods include, but are not limited to:

1. Rolling more or fewer dice for a roll or two 
2. Throwing the dice in an unusual way
3. Changing which country to attack after a series of bad rolls, in order to "reset" the dice
4. Screaming loudly to shock the dice into rolling 6s, or acting very subdued to not disturb them

While I can't speak to the efficacy of the last 3, I was quite excited to realize that there is an easy way to evaluate the first: simulate it! By the [Law of Large Numbers](https://en.wikipedia.org/wiki/Law_of_large_numbers), we can estimate the probabilites of success for each combination of number of attacking and defending dice. 

## Setting up our simulation

The first thing we need to do is recognize the algorithm we follow whenever we roll in Risk. After declaring the country to attack and the origin of the attack: 

1. Both the attacker and the defender pick the number of dice they would like to roll. The defender can roll up to 2, and the attacker can roll up to 3.
2. Both players roll. 
3. After the roll, they (visually) sort their numbers from highest to lowest.
4. The highest number of each player is compared against the other. Whoever's number is highest wins, and one of the opposing player's armies is removed from the board. The defender wins ties.
5. If both of the players rolled at least two dice, the second highest number is also compared, in the same way as in step 4. 

Each of these steps is easy to complete in R. For example, we can wrap the `sample` function to simulate a dice roll.

```{r rolling}
roll <- function(dice = 1) {
  sample(1:6, size = dice, replace = TRUE)
}

t(replicate(4, roll(dice = 2)))
```

The rest of the code below sets up the other functions that we need. The key function is `battle`, which uses the other functions to implement this algorithm.

```{r}
# Get the top roll(s) for a player. We only care about the top n rolls, where n 
# is either 1 or 2, and represents min(num_rolled_by_attacker,
# num_rolled_by_defender)
get_top <- function(roll, n) {
  num <- min(length(roll), n)
  sorted <- sort(roll, decreasing = TRUE)
  
  if(num > 2 | num <= 0) {
    stop("invalid number of dice rolled - must be 1, 2, or 3 for attacker and 1 or 2 for defender.")
  }
  else if (num == 1) {
    return(sorted[1])
  }
  else {
    return(c(sorted[1], sorted[2]))
  }
}

# compare each pair of dice. Defense wins ties.
compare <- function(attack, defense) {
  if(attack > defense){
    return("attack")
  }
  else return("defense")
}

# one dice roll from both defense and offense. Resulting vector will indicate
# who won each comparison
battle <- function(num_d = 2, num_o = 3) {
  if(num_d < 1 | num_d > 2) {
    stop("invalid number of defensive dice rolled: must be 1 or 2")
  }
  else if(num_o < 1 | num_o > 3) {
    stop("invalid number of offensive dice rolled: must be 1, 2, or 3")
  }
  d <- roll(num_d)
  o <- roll(num_o)
  min_dice <- min(num_d, num_o)
  
  d_top <- get_top(d, min_dice)  
  o_top <- get_top(o, min_dice)
  
  if(min(num_d, num_o) == 1){
    obv <- compare(o_top[1], d_top[1])
  }
  else {
    obv <- c(compare(o_top[1], d_top[1]), compare(o_top[2], d_top[2]))
  }
  
  obv
}

# an example of the output we can expect:
battle()

```

## Running the simulation and results

We can now run our simulation. To get an accurate result, we'll have to call our `battle` function a large number of times for each set of potential inputs. Because we can, we'll run it 100,000 times for each input permutation. To ease this, and to calculate our ultimate probabilities, we'll use a few helper functions:

```{r}
sim <- function(n, num_d, num_o) {
  t(replicate(n, battle(num_d = num_d, num_o = num_o)))
}

# probability the attacker wins
prob_attacker <- function(result) {
  length(result[result == "attack"]) / length(result)
}

prob_defender <- function(prob_attacker) {
  return(1 - prob_attacker)
}


```

And now, for our calculation of our actual results. We'll build two matrices: the row index will indicate how many dice the defender rolls, while the column index will indicate the same for the attacker. 

```{r}
attacker_likelihood <- matrix(data = NA, nrow = 2, ncol = 3)

for(i in 1:2) {
  for (j in 1:3) {
    attacker_likelihood[i, j] <- prob_attacker(sim(n = 100000, num_d = i, num_o = j))
  }
}

```
This takes ~30-45 seconds to run on my computer. Here are the results that I found, looking at the probability of success on the **attacker's** side:

```{r pretty-fying, echo = FALSE}
library(pander)

colnames(attacker_likelihood) <- c("Attack rolls 1", "Attack rolls 2", "Attack rolls 3")
rownames(attacker_likelihood) <- c("Defense rolls 1", "Defense rolls 2")

pandoc.table(attacker_likelihood)

```

And on the **defender's** side:

```{r pretty-fying-for-defender, echo = FALSE}
defender_likelihood <- prob_defender(attacker_likelihood)

pandoc.table(defender_likelihood)

```

I was interested to see that in the classic "showdown" setting (attacker rolls 3, defender rolls 2), the attacker has a slight advantage (54%). In general, however, there's one clear message from these numbers: always roll as many dice as you can in Risk!

(Note: If you're interested in a deeper analysis of Risk, you might want to check out [this article](http://www.datagenetics.com/blog/november22011/).)
